---
title: "An Exhaustive Analysis of ALL Great Energy Predictor Variables"
author: "Naquib Alam"
date: "09/05/2021"
output:
  html_document:
    code_folding: hide
    theme: journal
    highlight: tango
    df_print: paged
    fig_height: 7
    fig_width: 10
    number_sections: no
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this competition, we are tasked with developing accurate predictions of metered building energy usage in the following areas: chilled water, electric, natural gas, hot water, and steam meters. The data comes from over 1,000 buildings over a three-year timeframe.

The aim is that with better estimates of these energy-saving investments, large scale investors and financial institutions will be more inclined to invest in this area to enable progress in building efficiencies.


# The Data

There are five files that have been provided and also a sample submission file.

The five primary files to be used are described below, with the variable names also included:


**train.csv**:

`building_id` - Foreign key for the building metadata.
`meter` - The meter id code. Read as {0: electricity, 1: chilledwater, 2: steam, hotwater: 3}. Not every building has all meter types.
`timestamp` - When the measurement was taken
`meter_reading` - The target variable. Energy consumption in kWh (or equivalent). Note that this is real data with measurement error, which we expect will impose a baseline level of modeling error.


**building_meta.csv**:

`site_id` - Foreign key for the weather files.
`building_id` - Foreign key for training.csv
`primary_use` - Indicator of the primary category of activities for the building based on EnergyStar property type definitions
`square_feet` - Gross floor area of the building
`year_built` - Year building was opened
`floor_count` - Number of floors of the building


**weather_[train/test].csv**:

Weather data from a meteorological station as close as possible to the site.

`site_id`
`air_temperature` - Degrees Celsius
`cloud_coverage` - Portion of the sky covered in clouds, in oktas
`dew_temperature` - Degrees Celsius
`precip_depth_1_hr` - Millimeters
`sea_level_pressure` - Millibar/hectopascals
`wind_direction` - Compass direction (0-360)
`wind_speed` - Meters per second


**test.csv**

The submission files use row numbers for ID codes in order to save space on the file uploads. test.csv has no feature data; it exists so you can get your predictions into the correct order.

`row_id` - Row id for your submission file
`building_id` - Building id code
`meter` - The meter id code
`timestamp` - Timestamps for the test data period



```{r, message=FALSE, warning=FALSE}
# load in libraries
library(tidyverse)
library(scales)
library(lubridate)
library(ggridges)


# set plotting theme baseline
theme_set(theme_minimal() +
            theme(axis.title.x = element_text(size = 15, hjust = 1),
                  axis.title.y = element_text(size = 15),
                  axis.text.x = element_text(size = 12),
                  axis.text.y = element_text(size = 12),
                  panel.grid.major = element_line(linetype = 2),
                  panel.grid.minor = element_line(linetype = 2),
                  plot.title = element_text(size = 18, colour = "grey25", face = "bold"), plot.subtitle = element_text(size = 16, colour = "grey44")))

col_pal <- c("#5EB296", "#4E9EBA", "#F29239", "#C2CE46", "#FF7A7F", "#4D4D4D")


# read in data
train <- data.table::fread("../input/ashrae-energy-prediction/train.csv")
# test <- read.csv("data/ashrae-energy-prediction/test.csv", stringsAsFactors = F)
# 
# # # add column to train and test to indicate which df they are when joined
# train$df <- "train"
# test$df <- "test"
# 
# train_test <- bind_rows(train, test)
# rm(train, test);gc()

building <- read.csv("../input/ashrae-energy-prediction//building_metadata.csv")


weather_train <- read.csv("../input/ashrae-energy-prediction/weather_train.csv")
# weather_test <- read.csv("data/ashrae-energy-prediction/weather_test.csv", stringsAsFactors = F)
# 
# weather <- bind_rows(weather_train, weather_test)
# rm(weather_train, weather_test);gc()

# samp_sub <- weather_train <- read.csv("data/ashrae-energy-prediction/sample_submission.csv", stringsAsFactors = F)


# # join all data into one complete file
# combined_df <- train_test %>%
#   left_join(building, by = "building_id") %>%
#   left_join(weather, by = c("site_id", "timestamp"))
# 
# rm(train_test, weather, building);gc()

# join all data into one complete file
combine_train <- train %>%
  left_join(building, by = "building_id") %>%
  left_join(weather_train, by = c("site_id", "timestamp"))


rm(train, building, weather_train)

```


# Feature Engineering

The code block below will be expanded on over time as I come up with some new features.

Engineered features include:

* Month of the year
* Day of the week of the timestamp
* Hour of the day

```{r, warning=FALSE, message=FALSE}
combine_train <- combine_train %>% 
  mutate(timestamp_date = ymd(gsub( " .*$", "", timestamp)),
         timestamp_month = month(timestamp_date),
         timestamp_day = wday(timestamp_date, label = T, abbr = T),
         timestamp_day_number = day(timestamp_date),
         time_ymd_hms = ymd_hms(timestamp),
         time_hour = hour(time_ymd_hms))

```



# Exploratory Data Analysis (EDA)

To conduct the following analysis, the `train`, `building` and `weather_train` datasets have been joined. I had tried combining the training and testing sets together, but found processing time for the following EDA to be slowed down considerably, so will replicate any cleaning / feature engineering conducted on the `train` to the `test` set.

The joined dataframe (`combine_train`) now has `r comma(nrow(combine_train))` rows, and `r ncol(combine_train)` columns.


## Missing Data

The missing data is shown below. The observations with missing weather and building variables will need to be investigated throughout the report


```{r, message=FALSE, warning=FALSE}
colSums(is.na(combine_train))
```



## Target Variable Analysis - `meter_reading`


```{r, message=FALSE, warning=FALSE}
combine_train %>% 
  ggplot(aes(x=meter_reading)) +
  geom_histogram(alpha = 0.5, fill = "#5EB296", colour = "#4D4D4D") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  ggtitle("METER READING VARIABLE IS HEAVILY SKEWED BY SOME OUTLIERS", subtitle = "We will remove those and visualise") +
  labs(x= "Meter Reading", y= "Count")

summary(combine_train$meter_reading)
```


To visualise the distribution of `meter_reading` without outliers, I will split those outliers used by the formula (1.5 x IQR).

As we can see, the target variable is still heavily skewed, even for those observations within the outlier cutoff.


```{r, message=FALSE, warning=FALSE}
outlier <- round(1.5 * IQR(combine_train$meter_reading),0)

combine_train %>% 
  mutate(outlier = ifelse(meter_reading > outlier, "Outlier", "Not Outlier")) %>% 
  ggplot(aes(x=meter_reading)) +
  geom_histogram(alpha = 0.5, fill = "#5EB296", colour = "#4D4D4D") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  ggtitle("METER READING VARIABLE IS HEAVILY SKEWED", subtitle = "Meter Reading still skewed even for non-outliers") +
  labs(x= "Meter Reading", y= "Count") +
  facet_wrap(~ outlier, scales = "free")
```


We will definitely be performing some transformations on `meter_reading` for modelling purposes - see below:

```{r, message=FALSE, warning=FALSE}
combine_train %>% 
  ggplot(aes(x= log(meter_reading + 1))) +
  geom_histogram(alpha = 0.5, fill = "#5EB296", colour = "#4D4D4D") +
  scale_y_continuous(labels = comma) +
  ggtitle("LOG TRANSFORMING METER READING VARIABLE IS NECESSARY", subtitle = "The variable looks a lot more workable now") +
  labs(x= "log(Meter Reading)", y= "Count")
  
```


### Meter Reading and Meter Type


There are four different meter types and are displayed below:

* 0: electricity
* 1: chilledwater 
* 2: steam
* 3: hotwater


We can see that the `steam` meter type tends to have higher meter readings, while `electricity` tends to have the lowest energy readings.


```{r, message=FALSE, warning=FALSE}
combine_train %>% 
  ggplot(aes(x= log(meter_reading + 1), fill = as.character(meter))) +
  geom_density(alpha = 0.5, adjust = 2) +
  scale_fill_manual(values = col_pal) +
  ggtitle("THE STEAM METER TYPE IS THE LEAST EFFICIENT, ELECTRICITY\nTHE MOST EFFICIENT", subtitle = "Meter readings of 0 have been excluded") +
  labs(x= "log(Meter Reading)") +
  theme(axis.title.y = element_blank(), legend.position = "top")
```


### Weekday and Meter Reading

I suppose it's to be expected, Sundays tend to generally have lower readings, but there isn't a major difference between the days


```{r, warning=FALSE, message=FALSE}
combine_train %>% 
  ggplot(aes(x= timestamp_day, y= log(meter_reading + 1))) +
  geom_boxplot(alpha = 0.5, fill = "#5EB296", colour = "#4D4D4D") +
  ggtitle("SUNDAYS HAVE THE LOWEST READINGS", subtitle = "Not much difference between the days though") +
  labs(y= "log(Meter Reading)") +
  coord_flip() +
  theme(axis.title.y = element_blank())
```


### Time of Day and Meter Reading

Reading are significantly higher during traditional work hours and this is to be expected. Time of day appears like it will be a significant predictor in any subsequent model for this competition.

```{r, warning=FALSE, message=FALSE}
combine_train %>% 
  group_by(time_hour) %>% 
  summarise(median_reading = median(meter_reading, na.rm = T)) %>% 
  ggplot(aes(x= time_hour, y= median_reading)) +
  geom_line(colour = "#5EB296", size = 1) +
  ggtitle("READINGS HIGHEST DURING THE MIDDLE OF THE DAY", subtitle = "The lowest point is around 4am") +
  labs(x= "Reading Hour", y= "Median Reading")
```


There are considerable differences between building types as to when meter readings are highest. Healthcare buildings peak slightly later in the day, while Housing peaks from about 7pm. Manufacturing has higher readings earlier than the average, while Utility peaks before 5am.

```{r, warning=FALSE, message=FALSE}
combine_train %>% 
  group_by(time_hour, primary_use) %>% 
  summarise(median_reading = median(meter_reading, na.rm = T)) %>% 
  ggplot(aes(x= time_hour, y= median_reading)) +
  geom_line(colour = "#5EB296", size = 1) +
  ggtitle("TIME READINGS ARE HIGHEST CHANGES BASED ON BUILDING TYPE", subtitle = "Residential, and Warehouse peak later in day, Utility earlier") +
  labs(x= "Reading Hour", y= "Median Reading") +
  facet_wrap(~ primary_use, scales = "free")
```



### Primary Use and Meter Reading

We can see that `Utility` and `Healthcare` places tend to have the highest readings, while Religious Worship places the least - they're no doubt frequented less often than the higher energy users.

```{r, message=FALSE, warning=FALSE}
combine_train %>% 
  ggplot(aes(x= as.character(primary_use), y= log(meter_reading + 1))) +
  geom_boxplot(alpha = 0.5, fill = "#5EB296", colour = "#4D4D4D") +
  ggtitle("UTILITIES AND HEALTHCARE HAVE THE HIGHEST READINGS", subtitle = "Religious worship places the lowest") +
  labs(y= "log(Meter Reading)") +
  coord_flip() +
  theme(axis.title.y = element_blank())
```


It's interesting to see that different building types have different energy usage on different days - religious places have higher readings on weekends compared to the average, while places of industry (businesses) have higher readings on weekdays.

```{r, message=FALSE, warning=FALSE}
combine_train %>% 
  ggplot(aes(x= timestamp_day, y= log(meter_reading + 1))) +
  geom_boxplot(alpha = 0.5, fill = "#5EB296", colour = "#4D4D4D") +
  ggtitle("PLACES OF INDUSTRY HIGHEST READINGS ON WEEKDAYS", subtitle = "Religious worship places have higher readings on weekends") +
  labs(y= "log(Meter Reading)") +
  coord_flip() +
  theme(axis.title.y = element_blank()) +
  facet_wrap(~ primary_use, scales = "free")
```



### Meter Readings over time

Interestingly, the median meter reading per reading began rising in May and peaked between July and October 2016.

```{r, message=FALSE, warning=FALSE}
combine_train %>% 
  group_by(timestamp_date) %>% 
  summarise(median_reading = median(meter_reading, na.rm = T)) %>% 
  ggplot(aes(x= timestamp_date, y= median_reading)) +
  geom_line(colour = "#5EB296") +
  geom_smooth(se = F, colour = col_pal[3]) +
  ggtitle("READINGS REALLY PEAKED FROM MAY TO OCTOBER", subtitle = "Readings are the median reading per each reading") +
  labs(x= "Reading Date", y= "Median Reading")

```


The trend holds for most of the different building types, with a few notable exceptions; *Manufacturing* dips during that peak period outlined above, while *Services*, *Technology*, *Utility* and *Warehouse* remained fairly constant over the year.

```{r, message=FALSE, warning=FALSE}
combine_train %>% 
  group_by(timestamp_date, primary_use) %>% 
  summarise(median_reading = median(meter_reading, na.rm = T)) %>% 
  ggplot(aes(x= timestamp_date, y= median_reading)) +
  geom_line(colour = "#5EB296") +
  geom_smooth(se = F, colour = col_pal[3]) +
  ggtitle("MANUFACTURING REALLY BUCKED THE GENERAL TREND", subtitle = "While some other business types remained fairly stable over the period") +
  labs(x= "Reading Date", y= "Median Reading") +
  facet_wrap(~ primary_use, scales = "free") +
  theme(axis.text.x = element_text(angle = 45))

```



## Correlation between `meter_reading` and Numeric Variables

Both the number of floors a building has and the size of the building have a fairly strong Pearson correlation with the `meter_reading`, ie, the bigger the building, the higher the reading. The year the building was built is also moderately strongle positively correlated - older buildings may have higher readings.

With a very weak negative Pearson correlation, `wind_speed`, `air_temperature` and `cloud_coverage` may be important in any model that gets built.

```{r, warning=FALSE, message=FALSE}
combine_train %>% 
  select(meter_reading, square_feet, year_built, floor_count, air_temperature, cloud_coverage, dew_temperature, precip_depth_1_hr, sea_level_pressure, wind_speed) %>% na.omit() %>% 
  as.matrix() %>% 
  cor() %>% as.data.frame() %>% 
  select(meter_reading) %>% mutate(vars = row.names(.)) %>% 
  filter(vars != "meter_reading") %>% 
  ggplot(aes(x= reorder(vars, meter_reading), y= meter_reading)) +
  geom_col(fill = "#5EB296", colour = col_pal[6]) +
  labs(x= "Variables", y= "Correlation with meter_reading") +
  ggtitle("SQUARE FEET AND FLOOR COUNT HIGHLY CORRELATED\nWITH METER READING", subtitle = "The year building built moderate positive correlation also") +
  coord_flip() +
  theme(panel.grid.major.y = element_blank())
```




## Visualising the Independent Variables

The independent variables (the variables we'll use to help us predict `meter_reading`) will be analysed in this section. Analysing these variables also will allow us to see their distributions and if there are any issues with the data (missing variables, skeweness, extreme outliers, etc).


### Primary Use

There are `r sum(is.na(combine_train$primary_use))` missing records in the `primary_use` variable.

*Education* building types are by far the most frequently read building types, while *Office* is a fair bit behind, but still almost double that of the third most read meters of *Entertainment* buildings.

There is a considerably long tail, with the bottom 11 being significantly lower.

```{r, warning=FALSE, message=FALSE}
combine_train %>% 
  count(primary_use) %>% 
  ggplot(aes(x= reorder(primary_use, n), y= n)) +
  geom_col(colour = col_pal[6], fill = col_pal[4]) +
  scale_y_continuous(labels = comma) +
  labs(y= "Number of Readings") +
  ggtitle("EDUCATION BUILDINGS HAVE BY FAR THE MOST READINGS", subtitle = "Mainly five building types dominate the data set") +
  theme(axis.title.y = element_blank()) +
  coord_flip()
```



### Square Feet

There are `r sum(is.na(combine_train$square_feet))` missing values in the `square_feet` variable.

The distribution of square feet is positively skewed, mith a median of `r comma(median(combine_train$square_feet))` feet and mean of `r comma(mean(combine_train$square_feet))` square feet.

```{r, warning=FALSE, message=FALSE}
combine_train %>% 
  ggplot(aes(x= square_feet)) +
  geom_density(colour = col_pal[6], fill = col_pal[4], adjust = 2, alpha = 0.7) +
  scale_x_continuous(labels = comma, name = "Building Square Feet") +
  ggtitle("BUILDING SIZE IS POSITIVELY SKEWED", subtitle = paste0("The median size is ", comma(median(combine_train$square_feet)), " square feet")) +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank())
```



### Meter

There are `r comma(sum(is.na(combine_train$meter)))` missing records in the `meter` variable.

We can see that Electricity is read by far the most frequently, with over 12m records, while `hotwater` the other end had 1.26m records recorded.

```{r, warning=FALSE, message=FALSE}
combine_train %>% 
  count(meter) %>% 
  ggplot(aes(x= meter, y= n)) +
  geom_col(colour = col_pal[6], fill = col_pal[4]) +
  scale_y_continuous(labels = comma, name = "Count of Readings") +
  ggtitle("ELECTRICITY THE MOST FREQUENT METER TYPE MEASURED", subtitle = "Big drop off to Chilledwater, while Hotwater distant last")
```


### Year Built

There are `r comma(sum(is.na(combine_train$year_built)))` missing records in the `year_built` variable.

These missing years are displayed below. *Services* building types do not have a year built recorded at all, while 11 of the 16 different types are missing over half.

```{r, warning=FALSE, message=FALSE}
combine_train %>% 
  group_by(primary_use) %>% 
  summarise(n = n(),
            n_missing = sum(is.na(year_built))) %>% 
  mutate(proportion_missing = n_missing / n) %>% 
  ggplot(aes(x= reorder(primary_use, proportion_missing), y= proportion_missing)) +
  geom_col(colour = col_pal[6], fill = col_pal[4]) +
  scale_y_continuous(labels = percent) +
  labs(y= "Proportion Missing") +
  ggtitle("SERVICES BUILDINGS HAVE NO RECORDED YEAR BUILT", subtitle = "11 of the 16 building types with over half missing year built record") +
  theme(axis.title.y = element_blank()) +
  coord_flip()
```


Of the recorded year built, 1976 was the most frequent, with 55 buildings built.

```{r, warning=FALSE, message=FALSE}
combine_train %>% 
  distinct(building_id, .keep_all = T) %>% 
  group_by(year_built) %>% 
  summarise(n_buildings = n()) %>%
  ggplot(aes(x= year_built, y= n_buildings)) +
  geom_line(colour = col_pal[4], size = 1) +
  scale_y_continuous(name = "Number Built", limits = c(0,100)) +
  labs(x= "Year Built")
```


### Floor Count

There are `r comma(sum(is.na(combine_train$floor_count)))` missing records in the `floor_count` variable. Of the `r length(unique(combine_train$building_id))` unique building IDs, `r combine_train %>% distinct(building_id, .keep_all = T) %>% summarise(n_missing = sum(is.na(floor_count))) %>% pull(n_missing)` don't have a floor count.

Of those that do, we can see that there aren't too many buildings with more than 10 floors, while the median number of floors is `r median(combine_train$floor_count, na.rm = T)`.

```{r, warning=FALSE, message=FALSE}
combine_train %>% 
  distinct(building_id, .keep_all = T) %>% 
  ggplot(aes(x= floor_count)) +
  geom_density(adjust = 2, colour = col_pal[6], fill = col_pal[4], alpha = 0.7) +
  labs(x= "Floor Count", y= "Density") +
  ggtitle("NOT MANY BUILDINGS WITH MORE THAN 10 FLOORS", subtitle = paste0("Median floors is ", median(combine_train$floor_count, na.rm = T), " floors"))
```


### Air Temperature

There are `r comma(sum(is.na(combine_train$air_temperature)))` missing `air_temperature` records in our combined training set, however as helpfully pointed out by PC Jimmy in the comments section, there are only 55 missing air temp records in the `weather_train` data set. I will create a function to fill these missing values.

Filling these missing values will not be a simple mean/median imputation, but rather a two step process:

1. Take the air_temperature value for the last timestamp as the next timestamps value
2. If the previous timestamp is also NA, then take the average of that day's temperature for that particular `site_id`

Following this process results in no further NAs in the `air_temperature` variable.

Conveniently, the function can also be applied to `dew_temperature` also using `dplyr::enquo`.

```{r, warning=FALSE, message=FALSE}
fill_missing_temp <- function(df, temp_var) {
  
  temp_var <- enquo(temp_var)
  
  df %>% 
    arrange(site_id, timestamp) %>% 
    group_by(site_id) %>% 
    mutate(last_temp = lag(!! temp_var)) %>% 
    ungroup() %>% 
    mutate(month_day = paste0(timestamp_month, "_", timestamp_day)) %>% 
    group_by(site_id, month_day) %>% 
    mutate(avg_daily_temp = mean(!! temp_var, na.rm = T)) %>%  
    ungroup() %>% 
    mutate(temp_clean = ifelse(is.na(!! temp_var), last_temp, !! temp_var),
           temp_clean = ifelse(is.na(temp_clean), avg_daily_temp, temp_clean)) %>% 
    select(-last_temp, -avg_daily_temp) %>% 
    pull(temp_clean)
  
} 


# now apply to the combined DF
combine_train$air_temp_clean <- fill_missing_temp(combine_train, air_temperature)
```

Plotting the variable shows the variable to be fairly normally distributed, with the majority of recordings being between ~13 and 25 degrees.

```{r, warning=FALSE, message=FALSE}
combine_train %>% 
  ggplot(aes(x= air_temp_clean)) +
  geom_density(adjust = 2, colour = col_pal[6], fill = col_pal[4], alpha = 0.7) +
  labs(x= "Air Temperature", y= "Density") +
  ggtitle("MOST TEMPERATURES RANGE BETWEEN 13-25 DEGREES", subtitle = paste0("Median air temperature is ", median(combine_train$air_temp_clean, na.rm = T), " degrees"))
```


### Dew Temperature

There are `r comma(sum(is.na(combine_train$dew_temperature)))` missing `dew_temperature` records in our combined training set. The function created above can be applied to fill these missing values.

```{r, warning=FALSE, message=FALSE}
combine_train$dew_temp_clean <- fill_missing_temp(combine_train, dew_temperature) 

combine_train %>% 
  ggplot(aes(x= dew_temp_clean)) +
  geom_density(adjust = 2, colour = col_pal[6], fill = col_pal[4], alpha = 0.7) +
  labs(x= "Dew Temperature", y= "Density") +
  ggtitle("MOST DEW TEMPERATURES RANGE BETWEEN 13-25 DEGREES", subtitle = paste0("Median dew temperature is ", median(combine_train$dew_temp_clean, na.rm = T), " degrees"))

```


**TBC**
